# CrossFit 2015 Leaderboard

### by Vitor Bernardes

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
library(tidyverse)
library(gridExtra)
theme_set(theme_light())
```

```{r echo=FALSE, Load_the_Data}
# Load the Data
cf <- read.csv('crossfit2015.csv')

# order 'howlong' variable
levels(cf$howlong) <- ordered(c('Less than 6 months|',
                                '6-12 months|',
                                '1-2 years|',
                                '2-4 years|',
                                '4+ years|'))
```

## Introduction to CrossFit and the Data Set

CrossFit is a popular fitness program and fitness sport created in 2000. It
combines elements of aerobic exercise, calisthenics (body weight exercises),
and Olympic weightlifting with the goal of improving overall fitness.

On the sport side, since 2007 CrossFit promotes an annual competition open for
athletes from all over the world, called the CrossFit Games. The Games has
three stages of qualification: the Open, Regionals, and the Games themselves.

The Open, which receives its name because participation is open to anyone, is
held over five weeks at the beginning of the competition season. Each week
contains a workout that must be completed by athletes. The athletes can complete
the workout at their local box (how CrossFit gyms are called) and submit their
scores online. The workouts are referenced by their year and the number
corresponding to the order they have been presented in. For example, the first
workout of the 2015 Open is called 15.1, the second one is called 15.2, and
so forth.

The data set we are going to analyze is the 2015 Open leaderboard. It contains
data from athletes from all over the world and the results they submitted for
each completed workout.

## Summary of the Data Set

Let's review the data set we are working with.

```{r echo=FALSE}
length(table(cf$athlete_id))
```

As we can see, it contains observations about roughly over 250,000 athletes
that competed in the 2015 Open.

Let's what data we have about each athlete.

```{r echo=FALSE}
str(cf)
```

We can see we have several variables with data on the athletes themselves (such
as name, region, age, height, and weight), some variables related to the Open
workouts and results (such as stage, category, score, and rank), and also some
results for benchmark workouts by each athlete (such as fran, helen, snatch,
and deadlift).

We will primarily be interested to see what factors are related to the athletes’
results, contained in the variables *score* and *rank*.

## Summary of Features

Let's briefly examine the features we will be using in our analysis in order to
identify their distribution, any outliers, and also improve our knowledge of the
data we will be working with.

```{r fig.width=9, echo=FALSE, message=FALSE, warning=FALSE, Feature_Summary}
# First, let's create a data frame with just one row per athlete, instead
# of one row per result, so we can plot their demographic data.
unique_athletes <- subset(cf, !duplicated(athlete_id))

division_bar <- qplot(x = division, data = unique_athletes)
category_bar <- qplot(x = category, data = unique_athletes)
age_hist <- qplot(x = age, data = unique_athletes)
height_hist <- qplot(x = height, data = unique_athletes)
weight_hist <- qplot(x = weight, data = unique_athletes)
snatch_hist <- qplot(x = snatch, data = unique_athletes)
deadlift_hist <- qplot(x = deadlift, data = unique_athletes)
pullups_hist <- qplot(x = pullups, data = unique_athletes)

grid.arrange(division_bar, category_bar, age_hist,
             height_hist, weight_hist, snatch_hist,
             deadlift_hist, pullups_hist, ncol = 3)
```

We can see there are some pretty extreme values for height, weight, snatch,
deadlift, and pullups that are getting in the way of our understanding the data.
Let's identify those outliers and remove them in order to make our analysis more
robust.

```{r fig.width=9, echo=FALSE, message=FALSE, warning=FALSE, Feature_Summary_No_Outliers}
is_outlier <- function(var) {
  var %in% boxplot.stats(var)$out
}

height_hist <- qplot(x = height,
                     data = subset(unique_athletes, !is_outlier(height)))
weight_hist <- qplot(x = weight,
                     data = subset(unique_athletes, !is_outlier(weight)))
snatch_hist <- qplot(x = snatch,
                     data = subset(unique_athletes, !is_outlier(snatch)))
deadlift_hist <- qplot(x = deadlift,
                     data = subset(unique_athletes, !is_outlier(deadlift)))
pullups_hist <- qplot(x = pullups,
                     data = subset(unique_athletes, !is_outlier(pullups)))

grid.arrange(division_bar, category_bar, age_hist,
             height_hist, weight_hist, snatch_hist,
             deadlift_hist, pullups_hist, ncol = 3)
```

Now this looks much better and provides us with a first look at the data we
will be working with and its distribution.

## Getting to know our athletes

Let's get to know a little about the athletes we will be exploring further in
this analysis.

### Age

```{r echo=FALSE, Athlete_Ages}
ggplot(data = subset(unique_athletes, !is.na(gender)),
       aes(x = age)) +
  geom_histogram(binwidth = 1, aes(fill = gender)) +
  scale_fill_brewer() +
  facet_wrap(~gender)
```

We see the distribution of the number of athletes by age is pretty similar for
both men and women. We can also see the number of male athletes is larger in
the 2015 Open.

### Category

The competition is divided into two categories: Rx and Scaled. In the Rx
category, athletes must complete the workouts exactly as prescribed. The Scaled
category was created so the Open would be more accessible to a larger
number of athletes, and has scaled-down versions of the Rx workouts.

Let’s see how the athletes are divided into both categories.

```{r echo=FALSE, Athlete_Categories}
ggplot(data = subset(unique_athletes, !is.na(gender)),
       aes(x = category, fill = division)) +
  geom_bar(position = 'dodge') +
  scale_fill_brewer(palette = 'Greens')
```

```{r echo=FALSE}
prop.table(table(unique_athletes$category))
```

This plot shows us the absolute majority of athletes (85%) in the 2015 Open are
in the Rx category. The plot also shows the proportion between male and female
athletes on both categories. While men are in higher number in the Rx category,
the Scaled category includes more women than men.

Now let's create a single plot where we will be able to see the distribution of
age per category.

```{r echo=FALSE, Athlete_Age_by_Categories}
ggplot(data = subset(unique_athletes, !is.na(gender)),
       aes(x = gender, y = age)) +
  geom_boxplot(aes(fill = category)) +
  scale_fill_brewer()
```

It is interesting to note the center of the distribution of scaled athletes
appears slightly higher than for Rx athletes. That is specially noticeable
for men. It seems reasonable, because older athletes might find it more
difficult to complete Rx workouts.

Let’s plot the histogram and some summary statistics for the scaled category to
check that observation.

```{r echo=FALSE, Scaled_Age_Histogram}
scaled_athletes <- subset(unique_athletes, category == 'Scaled')

ggplot(data = subset(scaled_athletes, !is.na(gender)),
       aes(x = age)) +
  geom_histogram(binwidth = 1, fill = 'lightskyblue3') +
  geom_vline(data = subset(scaled_athletes, gender == 'Female'),
             aes(xintercept = mean(subset(scaled_athletes,
                                          gender == 'Female')$age)),
             linetype = 2) +
  geom_vline(data = subset(scaled_athletes, gender == 'Male'),
             aes(xintercept = mean(subset(scaled_athletes,
                                          gender == 'Male')$age)),
             linetype = 2) +
  facet_wrap(~gender)
```

```{r echo=FALSE, Scaled_Age_Stats}
summary(subset(unique_athletes, gender == 'Female' & category == 'Rx')$age)
summary(subset(unique_athletes, gender == 'Male' & category == 'Rx')$age)

summary(subset(unique_athletes, gender == 'Female' & category == 'Scaled')$age)
summary(subset(unique_athletes, gender == 'Male' & category == 'Scaled')$age)
```

Indeed, we can see that while the average age for men and women in the Rx
category is very close, the Scaled category has higher average ages for both
genders.

### CrossFit experience

Let's find out how long the athletes have had CrossFit experience prior to
joining the Open.

```{r echo=FALSE, How_Long_Athletes}
ggplot(subset(unique_athletes, !is.na(howlong)),
       aes(x = howlong)) +
  geom_bar(fill = 'lightblue')
```

While we do not have data on many athletes, we can see many of them have
joined the 2015 Open with less than a year of CrossFit experience, which might
show eagerness to participate in the event.

One aspect that might influence the choice of Scaled vs. Rx category is how
long the athlete has been practicing CrossFit for. It seems reasonable that
more experienced athletes might be more inclined to opt for the Rx category.

```{r echo=FALSE, Category_by_How_Long}
ggplot(subset(unique_athletes, !is.na(howlong)),
       aes(x = howlong)) +
  geom_bar(aes(fill = category),
           position = 'dodge')
```

We can see the proportion of Rx athletes is larger for athletes with over 2
years of experience, and smaller for athletes with between 6 months and 2 years
of experience. One interesting observation is that most athletes with less than
6 months of CrossFit experience chose the Rx category. It certainly is a curious
fact to notice, however since we unfortunately don’t have experience data for
many of our athletes, we can’t draw many conclusions from it.

### Regions

Finally, let's take a look at where our athletes come from.

```{r echo=FALSE, Athlete_Regions}
ggplot(data = subset(unique_athletes, region != ""),
       aes(x = region)) +
  geom_bar(fill = 'lightblue') +
  coord_flip()
```

This plot shows that, despite being open to participation to athletes from all
over the world, the popularity of the competition is still heavily centered in
North America, followed by Europe. Huge continents such as Africa and Asia still
show very little participation in the Open.

## Taking a look at the workouts

Now let's take a look at some results.

As we mentioned, the workouts are referenced by their year and number, such as
15.1, 15.2 etc. We will refer to them using this pattern.

*NOTE: The 2015 Open had a special workout on the first week, which we will*
*refer to as 15.1a. In our data set, it is refered to as 1.1.*

Also, it is important to mention the workouts can be one of two kinds. In the
first kind, the athlete tries to achieve the highest possible number of
repetitions, or reps, in the given timeframe. In the second kind, the athlete
must complete the workout as fast as possible. The scores for the first kind of
workout are measured in number of reps (which means the higher, the better),
and for the second kind are measured in seconds (which means the lower, the
better).

All but the last workout of the 2015 Open are of the first kind. Only the 15.5
workout score is measured in seconds.

Now let's plot the distribuition of scores by workout and division. Unless
otherwise mentioned, we will focus on the Rx category for the analysis.

```{r echo=FALSE, Scores_by_Workout_and_Division}
ggplot(subset(cf, !is.na(cf$score) & category == 'Rx'),
       aes(x = score)) +
  geom_histogram(binwidth = 5, aes(fill = division)) +
  facet_wrap(c('stage', 'division'), scales = 'free')
```

Several plots show peaks. The peaks are present on 15.1, 15.2, and 15.4, but
they are particularly sharp and intriguing on workout 15.3. I should investigate
further to find out what happended there.

That is a very interesting chart, but warrants a closer look at each workout
so we can better understand the story they are telling.

### Workout 15.1

Workout 15.1 consisted of:

*Complete as many rounds and reps as possible in 9 minutes of:*
*15 toes-to-bars*
*10 deadlifts (115 / 75 lb.)*
*5 snatches (115 / 75 lb.)*

```{r echo=FALSE, message=FALSE, warning=FALSE, '15_1_Scores'}
open.15.1.rx <- subset(cf, category == 'Rx' & stage == 1)

ggplot(open.15.1.rx, 
       aes(x = score)) +
  geom_histogram(binwidth = 5, aes(fill = division)) +
  facet_wrap(~division, scales = 'free') +
  scale_x_continuous(breaks = seq(0, 300, 30))
```

Each 30 reps represents a completed round of exercises. This particular workout
is interesting as the first movement is a relatively easier gymnastic one,
compared to the other 2 weightlifting exercises. So the peaks in this plot show
how many people struggled to perform the weightlifting exercises in each round.
The dips at 15, 45, 75, and so on, show that athletes rarely ended their
workouts on the gymnastic movement, but rather on the heavier exercises.

Since the last 2 exercises for each round are the deadlift and the snatch, and
our data set contains benchmarks for those exercises for some athletes, let's
check if their maximum weight lifted had any relationship to their results in
this workout.

```{r echo=FALSE, message=FALSE, warning=FALSE, '15_1_Scores_by_Deadlift'}
ggplot(open.15.1.rx,
       aes(x = deadlift, y = score)) +
  geom_jitter(alpha = 1/100, aes(color = division)) +
  geom_smooth(data = subset(open.15.1.rx, division == 'Male'),
              color = 'blue', linetype = 2) +
  geom_smooth(data = subset(open.15.1.rx, division == 'Female'),
              color = 'red', linetype = 2) +
  scale_color_brewer(palette = 'Set1') +
  scale_x_continuous(limits=c(quantile(unique_athletes$deadlift, .01,na.rm=T),
                              quantile(unique_athletes$deadlift, .99,na.rm=T)))+
  scale_y_continuous(limits=c(quantile(open.15.1.rx$score, .01, na.rm=T),
                              quantile(open.15.1.rx$score, .99, na.rm=T)))
```

Even though the data is very dispersed, we can see a positive trend between the
athlete’s record deadlift and their score on this workout, both for male and
female athletes.

Now let's run the same analysis for the snatch.

```{r echo=FALSE, message=FALSE, warning=FALSE, '15_1_Scores_by_Snatch'}
ggplot(data = open.15.1.rx,
       aes(x = snatch, y = score)) +
  geom_jitter(alpha = 1/100, aes(color = division)) +
  geom_smooth(data = subset(open.15.1.rx, division == 'Male'),
              color = 'blue', linetype = 2) +
  geom_smooth(data = subset(open.15.1.rx, division == 'Female'),
              color = 'red', linetype = 2) +
  scale_color_brewer(palette = 'Set1') +
  scale_x_continuous(limits=c(quantile(unique_athletes$snatch, .01,na.rm = T),
                              quantile(unique_athletes$snatch, .99,na.rm = T)))+
  scale_y_continuous(limits =c(quantile(open.15.1.rx$score, .05, na.rm = T),
                               quantile(open.15.1.rx$score, .99, na.rm = T)))
```

The same can be said for the snatch. Though the data is also very dispersed, we
can see a positive trend between the athlete’s record snatch and their score on
this workout.

### Workout 15.1a

Workout 15.1a consisted of:

*1-rep-max clean and jerk*
*6-minute time cap*

In other words, the athlete had 6 minutes to perform the heaviest clean and
jerk he or she could manage. This is a workout where strength is critical.

Let's see the distribution of scores for this workout.

```{r echo=FALSE, message=FALSE, warning=FALSE, '15_1a_Scores'}
open.15.1a.rx <- subset(cf, category == 'Rx' & stage == 1.1)

ggplot(open.15.1a.rx,
       aes(x = score)) +
  geom_histogram(binwidth = 10, aes(fill = division)) +
  facet_wrap(~division, scales = 'free')
```

The distribution of scores looks very similar for both divisions.

Since strength is vital for this workout, I wonder if bodyweight has any
relation to the score. Let's make a plot of result by weight and find out.

```{r echo=FALSE, message=FALSE, warning=FALSE, '15_1a_Scores_by_Weight'}
ggplot(data = open.15.1a.rx,
       aes(x = 5*round(weight/5), y = score)) +
  geom_jitter(aes(color = division),
              alpha = 1/170) +
  geom_line(stat = 'summary',
            fun.y = mean,
            aes(color = division),
            size = 1) +
  scale_x_continuous(limits=c(quantile(unique_athletes$weight, .01,na.rm = T),
                              quantile(unique_athletes$weight, .99,na.rm = T)))+
  scale_y_continuous(limits = c(50, 350))
```

```{r echo=FALSE}
cor.test(open.15.1a.rx$weight, open.15.1a.rx$score)
```

Indeed we see a positive relationship between bodyweight and score for 15.1a.

### Workout 15.2

Workout 15.2 consisted of:

*Every 3 minutes for as long as possible complete:*
*From 0:00-3:00*
*2 rounds of:*
*10 overhead squats (95 / 65 lb.)*
*10 chest-to-bar pull-ups*
*From 3:00-6:00*
*2 rounds of:*
*12 overhead squats (95 / 65 lb.)*
*12 chest-to-bar pull-ups*
*Etc., following same pattern until you fail to complete both rounds*

```{r echo=FALSE, message=FALSE, warning=FALSE, '15_2_Scores'}
open.15.2.rx <- subset(cf, category == 'Rx' & stage == 2)

ggplot(open.15.2.rx,
       aes(x = score)) +
  geom_histogram(binwidth = 5, aes(fill = division)) +
  facet_wrap(~division, scales = 'free')
```

Let's zoom in on the female plot and try to figure out why there is a sharp
peak at around score 10.

```{r echo=FALSE, message=FALSE, warning=FALSE, '15_2_Female_Scores_Zoom'}
ggplot(subset(open.15.2.rx, division == 'Female'), 
       aes(x = score)) +
  geom_histogram(binwidth = 1, fill = 'lightcoral') +
  scale_x_continuous(limits = c(0, 100), breaks = seq(0, 100, 5))
```

```{r echo=FALSE}
female.15.2.scores <- table(subset(open.15.2.rx, division == 'Female')$score)
female.15.2.scores[names(female.15.2.scores) == 10]

prop.female.15.2.scores <- prop.table(female.15.2.scores)
prop.female.15.2.scores[names(prop.female.15.2.scores) == 10]
```

Now we can see very clearly what happened: almost 9,500 women, or 17.3% of the
total number of athletes that completed this workout, were not able to complete
1 chest-to-bar pull-up after the first set of 10 overhead squats. The ones that
were able to perform that movement were able to pass the first round and
completed the workout at several score levels. That shows how difficult that
movement is to perform.

Our data set also contains data about the maximum number of pull-ups each
athlete has performed in a row. Let's plot the average 15.2 score by maximum
pull-ups.

```{r echo=FALSE, message=FALSE, warning=FALSE, '15_2_Mean_Scores_by_Pullup'}
ggplot(data = open.15.2.rx,
       aes(x = pullups, y = score)) +
  geom_jitter(aes(color = division),
              alpha = 1/30) +
  geom_line(stat = 'summary',
            fun.y = mean,
            aes(color = division),
            size = 1) +
  scale_x_continuous(limits=c(quantile(unique_athletes$pullups,.02,na.rm = T),
                              quantile(unique_athletes$pullups,.98,na.rm = T)))+
  scale_y_continuous(limits = c(0, 300))
```

This plot clearly shows a positive relationship between the average maximum
number of pull-ups and 15.2 scores. In addition, since the female score
by average maximum pull-ups is higher than male scores, it reinforces how
particularly challenging the chest-to-bar pull-up was for female athletes.

### Workout 15.3

Workout 15.3 consisted of:

*Complete as many rounds and reps as possible in 14 minutes of:*
*7 muscle-ups*
*50 wall-ball shots*
*100 double-unders*

```{r echo=FALSE, message=FALSE, warning=FALSE, '15_3_Scores'}
open.15.3.rx <- subset(cf, category == 'Rx' & stage == 3)

ggplot(open.15.3.rx,
       aes(x = score)) +
  geom_histogram(binwidth = 10, aes(fill = division)) +
  facet_wrap(~division, scales = 'free')
```

This is the plot we saw earlier that shows very sharp peaks. Let's dig deeper
and take a look at this workout to try to figure out what the peaks are about.
We will plot the women's division and look more closely at what happened.

```{r echo=FALSE, message=FALSE, warning=FALSE, '15_3_Female_Scores'}
ggplot(subset(open.15.3.rx, division == 'Female'),
       aes(x = score)) +
  geom_histogram(binwidth = 3, fill = 'lightcoral') +
  scale_x_continuous(limits = c(0, 700), breaks = seq(0, 700, 50))
```

This workout consisted of 3 different movements, the first of which was 7 reps
of a highly demanding movement (muscle-up), followed by 50 reps and 100 reps of
two much simpler movements. So this plot depicts a very clear story: athletes
struggling to perform the first movement, then speeding somewhat more easily
through the next two, and then struggling (again) if they are able to reach the
next round of muscle-ups. Each peak corresponds to athletes whose workouts ended
trying to execute that movement.

After the start of rounds two and three (second and third peaks), we can see a
smaller concentration of scores, which represent athletes whose workouts ended
executing the second movement in the round, the wall balls.

### Workout 15.4

Workout 15.4 consisted of:

*Complete as many reps as possible in 8 minutes of:*
*3 handstand push-ups*
*3 cleans*
*6 handstand push-ups*
*3 cleans*
*9 handstand push-ups*
*3 cleans*
*12 handstand push-ups*
*6 cleans*
*15 handstand push-ups*
*Etc., adding 3 reps to the handstand push-up each round, and 3 reps to the*
*clean every 3 rounds.*
*Men clean 185 lb.*
*Women clean 125 lb.*

Since this workout involves handstand pushups, which are pushups where the
athlete is upside down, pushing up his or her own weight, let’s investigate if
lighter athletes had an edge.

```{r echo=FALSE, message=FALSE, warning=FALSE, '15_4_Mean_Scores_by_Weight'}
open.15.4.rx <- subset(cf, category == 'Rx' & stage == 4)

ggplot(data = open.15.4.rx,
       aes(x = 5*round(weight/5), y = score)) +
  geom_jitter(aes(color = division),
              alpha = 1/100) +
  geom_line(stat = 'summary',
            fun.y = mean,
            aes(color = division),
            size = 1) +
  scale_x_continuous(limits=c(quantile(unique_athletes$weight, .01,na.rm = T),
                              quantile(unique_athletes$weight, .99,na.rm = T)))+
  scale_y_continuous(limits = c(0, 150))
```

That doesn’t look to be the case. We can identify, especially on the male
division, a curve where lighter and heavier athletes performed more poorly. We
should take this information and remember this workout also included a
weightlifting movement, so it makes sense that an optimal range of weights would
yield better results.

How about height? I wonder how it relates to scores for this workout.

```{r echo=FALSE, message=FALSE, warning=FALSE, '15_4_Mean_Scores_by_Height'}
ggplot(open.15.4.rx,
       aes(x = height, y = score)) +
  geom_jitter(aes(color = division),
              alpha = 1/100) +
  geom_line(stat = 'summary',
            fun.y = mean,
            aes(color = division),
            size = 1) +
  scale_x_continuous(limits = c(62, 77)) +
  scale_y_continuous(limits = c(0, 150))
```

This plot shows us something interesting and somewhat unexpected. We can
definitely see a trend where shorter athletes had better results on average for
this workout.

Now let's combine the two measurements and display both on one chart.

```{r echo=FALSE, message=FALSE, warning=FALSE, '15_4_M_Scores_Weight_Height'}
ggplot(subset(open.15.4.rx,
              height > 62 & height <= 75 & division == 'Male'),
       aes(x = 5*round(weight/5), y = score)) +
  geom_jitter(alpha = 1/5, aes(color = height)) +
  geom_smooth() +
  scale_color_gradient(low = 'white', high = 'red') +
  scale_x_continuous(limits = c(125, 275)) +
  scale_y_continuous(limits = c(0, 150)) +
  theme_dark()
```

```{r echo=FALSE, message=FALSE, warning=FALSE, '15_4_F_Scores_Weight_Height'}
ggplot(subset(open.15.4.rx,
              height >= 60 & height <= 70 & division == 'Female'),
       aes(x = 5*round(weight/5), y = score)) +
  geom_jitter(alpha = 1/5, aes(color = height)) +
  geom_smooth() +
  scale_color_gradient(low = 'white', high = 'red') +
  scale_x_continuous(limits = c(100, 180)) +
  scale_y_continuous(limits = c(0, 170)) +
  theme_dark()
```

On both the above charts we can see that lighter athletes tend to be shorter,
and the heavier athletes are usually taller. We can also see how the top of the
charts are lighter (indicating shorter athletes), especially on the lighter
bodyweight range.

### Workout 15.5

Workout 15.5 consisted of:

*27-21-15-9 reps for time of:*
*Row (calories)*
*Thrusters*
*Men use 95 lb.*
*Women use 65 lb.*

This looks like a pretty intensive workout. It involved completing the sequence
of movements as fast as possible, so the lower the score, the better the result
for the athlete. I wonder how age affected athletes’ performance.

```{r echo=FALSE, message=FALSE, warning=FALSE, '15_5_Score_by_Age'}
open.15.5 <- subset(cf, stage == 5 & category == 'Rx')

ggplot(open.15.5,
       aes(x = age, y = score)) +
  geom_jitter(alpha = 1/60, aes(color = division)) +
  geom_smooth() +
  scale_color_brewer(palette = 'Set1') +
  scale_x_continuous(limits = c(18, 55),
                     breaks = seq(10, 60, 5)) +
  scale_y_continuous(limits = c(300, 1500)) +
  facet_wrap(~division, scales = 'free')
```

Even though the points are very disperse, it looks like the older the athlete,
the longer he or she will take to complete this workout. Also, male athletes
appear to have achieved better scores on average than female athletes for this
workout.

Let's average the results by age to help us check this trend.

```{r echo=FALSE, message=FALSE, warning=FALSE, '15_5_Mean_Score_by_Age'}
ggplot(open.15.5,
       aes(x = age, y = score)) +
  scale_x_continuous(limits = c(18, 55),
                     breaks = seq(15, 60, 5)) +
  geom_jitter(aes(color = division),
              alpha = 1/100) +
  geom_line(stat = 'summary',
            fun.y = mean,
            aes(color = division),
            size = 1) +
  scale_y_continuous(limits = c(350, 1400))
```

The trend looks very clear now. For 15.5, average performance decreased with
age.

To continue with this reasoning, let's check if age is reflected on other
workouts as well. Let's plot the average result by age for the rest of the
workouts.

```{r fig.width=7, fig.height=10, echo=FALSE, message=FALSE, warning=FALSE, Mean_Scores_by_Age}
# Plot score by age for given stage
plot_score_by_age <- function(st, alp = 1/150) {
  plot <- ggplot(subset(cf,
                        category == 'Rx' &
                          stage == st),
                 aes(x = age, y = score)) +
  scale_x_continuous(limits = c(18, 55),
                     breaks = seq(15, 60, 5)) +
  geom_jitter(aes(color = division),
              alpha = alp) +
  geom_line(stat = 'summary',
            fun.y = mean,
            aes(color = division),
            size = 1)
  
  return(plot)
}

p1 <- plot_score_by_age(1) + scale_y_continuous(limits = c(25, 230))
p1a <- plot_score_by_age(1.1, alp = 1/170) +
  scale_y_continuous(limits = c(50, 350))
p2 <- plot_score_by_age(2) + scale_y_continuous(limits = c(0, 300))
p3 <- plot_score_by_age(3, alp = 1/80) + scale_y_continuous(limits = c(0, 500))
p4 <- plot_score_by_age(4) + scale_y_continuous(limits = c(0, 150))
p5 <- plot_score_by_age(5) + scale_y_continuous(limits = c(350, 1400))

grid.arrange(p1, p1a, p2, p3, p4, p5, ncol = 2)
```

Indeed, for most workouts, performance seems to peak at the age range of 20 to
30, then it starts to decrease from that age on. One interesting thing this
chart shows is the female performance on 15.3, which doesn't display that
trend as clearly, and shows relatively high mean scores for ages above 50.

Finally, let’s create a variable that represents each athlete’s overall rank and
plot that against age.

```{r echo=FALSE, message=FALSE, warning=FALSE, Create_Overall_Rank}
# The overall rank is calculated by summing the rank for each workout
overallrank_male <- cf[cf$category == 'Rx' & cf$division == 'Male', ] %>%
  group_by(athlete_id) %>%
  summarise(name = first(name),
            point_total = sum(rank))
overallrank_male$overall_rank <- rank(overallrank_male$point_total)
overallrank_male$overall_rank <- as.integer(overallrank_male$overall_rank)

overallrank_female <- cf[cf$category == 'Rx' & cf$division == 'Female', ] %>%
  group_by(athlete_id) %>%
  summarise(name = first(name),
            point_total = sum(rank))
overallrank_female$overall_rank <- rank(overallrank_female$point_total)
overallrank_female$overall_rank <- as.integer(overallrank_female$overall_rank)

# Join rank on main table
cf_rank <- cf %>%
  left_join(overallrank_male[, c('athlete_id', 'overall_rank')],
            by='athlete_id')
cf_rank <- cf_rank %>%
  left_join(overallrank_female[, c('athlete_id', 'overall_rank')],
            by='athlete_id')
cf_rank$overall_rank.x <- ifelse(!is.na(cf_rank$overall_rank.y),
                                 cf_rank$overall_rank.y,
                                 cf_rank$overall_rank.x)

# Keep just one column
cf_rank$overall_rank.y <- NULL
names(cf_rank)[names(cf_rank) == 'overall_rank.x'] <- 'overall_rank'
cf <- cf_rank
``` 

```{r echo=FALSE, message=FALSE, warning=FALSE, Overall_Rank_by_Age}
ggplot(subset(cf, category == 'Rx'),
       aes(x = age, y = overall_rank)) +
  scale_color_brewer(palette = 'Set1') +
  scale_x_continuous(limits = c(18, 55),
                     breaks = seq(10, 60, 5)) +
  geom_jitter(aes(color = division), alpha = 1/200) +
  geom_line(stat = 'summary',
            fun.y = mean) +
  facet_wrap(~division, scales = 'free')
```

We can see here the same trend we saw on the last chart. The average overall
rank by age seems to peak at 20 to 30, then it starts to decrease from that age
on.

## Final Plots and Summary

### Plot One
```{r echo=FALSE, warning=FALSE, Plot_One}
#15.2_Female_Scores_Zoom
ggplot(subset(open.15.2.rx, division == 'Female'), 
       aes(x = score)) +
  geom_histogram(binwidth = 1, fill = 'lightcoral') +
  scale_x_continuous(limits = c(0, 100), breaks = seq(0, 100, 5)) +
  scale_y_continuous(breaks = seq(0, 10000, 1000)) +
  labs(x = 'Score (Number of Repetitions)',
       y = 'Number of Athletes',
       title = 'Female 15.2 Scores - The Chest-to-Bar Obstacle')
```

This is a very simple — though impressive — chart, because it tells a story so
clearly.

The first round of this workout involved executing a set of 10 overhead squats,
followed by a set of 10 chest-to-bar pull-ups. This chart shows us the large
number of athletes who were able to complete the first 10 overhead squats, but
were unable to perform a single chest-to-bar pull-up. The number of athletes who
scored 10 repetitions on this workout was almost 9,500, or 17.3% of the total
number of competitors.

In other words, only by mastering that movement, athletes would have been
immediately ahead of 9,500 competitors.

### Plot Two
```{r echo=FALSE, warning=FALSE, message=FALSE, Plot_Two}
#15.4_Female_Scores_by_Weight_and_Height
ggplot(subset(open.15.4.rx,
              height >= 60 & height <= 70 & division == 'Female'),
       aes(x = 5*round(weight/5), y = score)) +
  geom_jitter(alpha = 1/5, aes(color = height)) +
  geom_smooth() +
  scale_color_gradient(low = 'white', high = 'red') +
  scale_x_continuous(limits = c(100, 180), breaks = seq(100, 180, 10)) +
  scale_y_continuous(limits = c(0, 170)) +
  theme_dark() +
  labs(x = 'Weight (pounds)',
       y = 'Score (Number of Repetitions)',
       color = 'Height (inches)',
       title = 'Female 15.4 Scores - Weight and Height')
```

This plot shows depicts the relationship between weight and scores for the
female division of Open 15.4, and also displays how heights are related to those
features.

Even though the scores are widely distributed across weights, averaging out
their values helps to identify a curve where the best average scores were
obtained by athletes weighing around 130 to 140 lbs.

Also, we can see that athletes’ weight tend to increase as height increases. In
addition, we can see that, as score increases, the horizontal “levels” get
slightly lighter, indicating a slightly better performance by shorter athletes
for this workout.

### Plot Three
```{r fig.width=7, fig.height=10, echo=FALSE, warning=FALSE, message=FALSE, Plot_Three}
p1 <- plot_score_by_age(1) +
  scale_y_continuous(limits = c(25, 230)) +
  ggtitle('15.1')
p1a <- plot_score_by_age(1.1, alp = 1/170) +
  scale_y_continuous(limits = c(50, 350)) +
  ggtitle('15.1a')
p2 <- plot_score_by_age(2) +
  scale_y_continuous(limits = c(0, 300)) +
  ggtitle('15.2')
p3 <- plot_score_by_age(3, alp = 1/80) +
  scale_y_continuous(limits = c(0, 500)) +
  ggtitle('15.3')
p4 <- plot_score_by_age(4) +
  scale_y_continuous(limits = c(0, 150)) +
  ggtitle('15.4')
p5 <- plot_score_by_age(5, alp = 1/100) +
  scale_y_continuous(limits = c(350, 1400)) +
  ggtitle('15.5')

set_labels <- function(plt) {
  plt <- plt + labs(x = 'Age (Years)',
                    y = 'Score (Number of Reps or Seconds)',
                    color = 'Division')
  return(plt)
}

p1 <- set_labels(p1)
p1a <- set_labels(p1a)
p2 <- set_labels(p2)
p3 <- set_labels(p3)
p4 <- set_labels(p4)
p5 <- set_labels(p5)

grid.arrange(p1, p1a, p2, p3, p4, p5,
             ncol = 2,
             top = 'Mean Scores by Age for Each Workout')
```


This is a somewhat depressing chart, but the trend it shows is very clear:
average CrossFit athletes’ performance decreases with age. Except for some high
scores for older female athletes on 15.3, we can clearly see that performance
peaks when the athlete is around 20 to 30 years old, then decreases from that
age on.

## Reflection

Exploring this data set enabled me to get a good sense of what relates to an
athlete’s performance at the Open. And, more often than not, it showed me what
is not related.

During the exploration, I hit several dead-ends, where expected relationships
between variables were very faint or appeared to be non-existent. It was only
after persisting and investigating multiple possibilities that some
relationships came to life.

The relationships that did appear were fascinating in two ways: they either
showed clear trends that marked an athlete’s performance, or they told precise
stories about what happened on a given week during the Open.

Regarding future work, I would love to incorporate the results from other years
and investigate how athletes’ profile and demographics changed over the years,
as well as the distribution of results. That would enable me to check if
athletes are getting better, and if the CrossFit Open is getting more
competitive each year.