---
title: "CrossFit 2015 Leaderboard"
author: "Vitor Sexto Bernardes"
date: "19/01/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

theme_set(theme_light())
```

## CrossFit 2015 Leaderboard

# Intro to data set

First we are going to explain what the data set is about.

# Show data set

Let's see a bit about the structure of the data we're working with.

```{r}
str(cf)
```

# First look at the data

## Getting to know our athletes

Let's see a little about the athletes contained in our data set.

Let's see their ages:

```{r athlete_ages}
# First, let's create a data frame with just one row per athlete, instead
# of one row per result, so we can plot their demographic data.
unique_athletes <- subset(cf, !duplicated(athlete_id))

ggplot(data = subset(unique_athletes, !is.na(gender)),
       aes(x = age)) +
  geom_histogram(binwidth = 1, aes(fill = gender)) +
  scale_fill_brewer() +
  facet_wrap(~gender)
```

Now let's see what categories they are in.

```{r athlete_categories}
ggplot(data = subset(unique_athletes, !is.na(gender)),
       aes(x = category, fill = division)) +
  geom_bar() +
  scale_fill_brewer(palette = 'Greens')
```

That's a big difference. Let's see the proportions in each division:

```{r}
prop.table(table(unique_athletes[unique_athletes$division == 'Female',]$category))
prop.table(table(unique_athletes[unique_athletes$division == 'Male',]$category))
```

TODO: Explain in percentages.

Also, let's plot them together!

```{r}
ggplot(data = subset(unique_athletes, !is.na(gender)),
       aes(x = age)) +
  geom_histogram(binwidth = 1, aes(fill = category)) +
  scale_fill_brewer() +
  facet_wrap(~gender)
```


Let's also see the regions they come from.

```{r athlete_regions}
ggplot(data = subset(unique_athletes, region != ""),
       aes(x = region)) +
  geom_bar(aes(fill = category)) +
  scale_fill_brewer(palette = 'Greens') +
  coord_flip()
```

Now let's divide the regions by gender as well, and see what we get.

```{r}
ggplot(data = subset(unique_athletes, !is.na(gender) & region != ''),
       aes(x = region)) +
  geom_bar(aes(fill = gender)) +
  scale_fill_brewer() +
  coord_flip()
```

## Taking a look at the workouts

Now let's take a look at some results.

Let's plot the distribuition of scores per workout and division. Unless
otherwise mentioned, we will focus on the Rx category.

```{r}
ggplot(subset(cf, !is.na(cf$score) & category == 'Rx'),
       aes(x = score)) +
  geom_histogram(binwidth = 5, aes(fill = division)) +
  facet_wrap(c('stage', 'division'), scales = 'free')
```

Things of note about the above chart:

- Most plots would benefit from a closer look, since many are long-tailed, due
to a few athletes with extreme scores standing out.
- I wonder what is up with those peaks at 15.3. I should take a closer look too.

### Workout 15.3

Let's dig deeper and take a look at this workout to try to figure out what
the peaks are about. We will plot the women's division for that.

```{r}
open.15.3 <- subset(cf, category == 'Rx' & stage == 3)

ggplot(subset(open.15.3, !is.na(cf$score) & division == 'Female'),
       aes(x = score)) +
  geom_histogram(binwidth = 5) +
  scale_x_continuous(limits = c(0, 700), breaks = seq(0, 700, 25))
```

Interesting. The workout consisted of 3 different exercises, the first of which
was 7 reps of a difficult movement (ring muscle-up), followed by 50 reps and
100 reps of two much simpler movements. So what we are seeing with this plot are
athletes struggling to perform the first movement, then speeding through the
next two, and then struggling (again) if they are able to reach the next round
of the first movement. Each peak corresponds to athletes whose workouts ended
trying to execute that movement.

Looking back at the other charts, we find that the peaks present on workouts
15.1 and 15.2 are also explained by the same thing. They consisted of several
movements, some of which were faster to execute than others.

## Relating results to factors

Now let's try to find some factors that might be correlated with the results
we see for different workouts.

### Workout 15.1a

Let's start by examining workout 15.1a. It consists of a basic 1 rep-max of a
clean & jerk movement. In other words, who is able to lift the largest weight
gets a better score. It seems reasonable to assume heavier athletes will have
a better chance at this workout than others. Let's check if that holds up.

```{r}
#open.15.1a <- subset(cf, stage == 1.1 & category == 'Rx')

ggplot(data = open.15.1a.rx,
       aes(x = 5*round(weight/5), y = score)) +
  geom_jitter(alpha = 1/100,
              color = 'lightblue3') +
  scale_color_continuous() +
  facet_wrap(~division, scales = 'free') +
  scale_x_continuous(limits = c(quantile(unique_athletes$weight, .005, na.rm = T),
                                quantile(unique_athletes$weight, .998, na.rm = T))) +
  geom_line(stat = 'summary',
            fun.y = mean,
            color = 'blue',
            linetype = 2)
```

Although the dispersion of weight lifted appears somewhat large, we can
definitively see an upward trend in average weight lifted from athletes with
around 100 to 200 lbs.

Let's take a closer look at just the averages then.

```{r}
ggplot(open.15.1a,
       aes(x = 5*round(weight/5), y = score)) +
  scale_x_continuous(limits = c(90, 300)) +
  geom_line(stat = 'summary',
            fun.y = mean,
            aes(color = division))
```

It looks like our assumption was correct: lighter athletes had a harder time on
this workout.

### Age

I wonder if age plays a significant role in the results.

First, let's take a look at the average overall rank by age.

```{r}
ggplot(subset(cf, category == 'Rx'),
       aes(x = age, y = overall_rank)) +
  scale_x_continuous(limits = c(18, 55),
                     breaks = seq(10, 60, 5)) +
  geom_line(stat = 'summary',
            fun.y = mean,
            aes(color = division))
```

That doesn't look like the relationship has any significance at all. Let's move
on and focus on the influence of age on a particularly intensive workout, 15.5.

### Age and 15.5

This workout involved completing the sequence of movements as fast as possible,
so the lower the score, the better the result for the athlete.

```{r}
open.15.5 <- subset(cf, stage == 5 & category == 'Rx')

ggplot(open.15.5,
       aes(x = age, y = score)) +
  geom_jitter(alpha = 1/40,
              aes(color = division)) +
  scale_color_brewer(palette = 'Set1') +
  scale_x_continuous(limits = c(18, 55),
                     breaks = seq(10, 60, 5)) +
  scale_y_continuous(limits = c(250, 1500))
```

The points are again very disperse. It looks like the older the athlete, the
longer he or she will take to complete this workout. Also, male athletes appear
to have achieved lower scores on average than female athletes for this workout.

Let's average the results by age to help us check this possible trend.

```{r}
ggplot(open.15.5,
       aes(x = age, y = score)) +
  scale_x_continuous(limits = c(18, 55),
                     breaks = seq(15, 60, 5)) +
  geom_line(stat = 'summary',
              fun.y = mean,
              aes(color = division))
```

### Age and other scores

To continue with this reasoning, let's check if age is reflected on other
workouts as well. Let's plot the average result by age for the rest of the
workouts.

```{r}
ggplot(subset(cf, category == 'Rx'),
       aes(x = age, y = score)) +
  scale_x_continuous(limits = c(18, 55),
                     breaks = seq(15, 60, 5)) +
  geom_line(stat = 'summary',
              fun.y = mean,
              aes(color = division)) +
  facet_wrap(~stage, scales = 'free')
```

Indeed, for most workouts, performance seems to peak at the age range of 20 to
30, then it starts to decrease from that age on.

One interesting thing this chart shows us is the female performance on 15.3,
which doesn't display that trend as clearly. Let's focus on that workout again
for a moment.

```{r}
# Top 10 results per age
top10.15.3.byage <- subset(open.15.3, division == 'Female') %>%
  group_by(age) %>%
  top_n(10, score)

ggplot(NULL,
       aes(x = age, y = score)) +
  scale_x_continuous(limits = c(18, 55),
                     breaks = seq(15, 60, 5)) +
  geom_line(data = subset(open.15.3, division == 'Female'),
            stat = 'summary',
            fun.y = mean,
            linetype = 2) +
  geom_point(data = top10.15.3.byage,
             aes(x = age, y = score)) +
  geom_line(data = top10.15.3.byage,
            aes(x = 3*round(age/3), y = score),
            stat = 'summary',
            fun.y = mean,
            linetype = 2) +
  scale_color_brewer()
```

### Height and scores

Similarly to age, let's check if height influences any workout result on
average.

```{r}
ggplot(subset(cf, category == 'Rx'),
       aes(x = height, y = score)) +
  scale_x_continuous(limits = c(60, 80)) +
  geom_line(stat = 'summary',
            fun.y = mean,
            aes(color = division)) +
  facet_wrap(~stage, scales = 'free')
```

It appears that, differently from age, height has no direct relationship to
workout scores.